Spatial and temporal variability in economic availability of residual biomass

### 2021-11-21 analysis/manuscript.Rmd
toast

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  bookdown::pdf_document2: default
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  bookdown::html_document2: default
header-includes:
- \usepackage[left]{lineno}
- \usepackage{setspace}
- \linenumbers
geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

@@

```{r libraries, echo=FALSE, message=FALSE}
setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra)
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
library(quantmod)
library(bookdown)
```

@@

Biofuels are becoming an increasingly important source of energy because
of their potential to reduce dependency on fossil fuels, thereby
reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
generation biofuel provides the benefit of avoiding competition with
food production by using lignocellulosic biomass -- often in the form of
agriculture and forest residues -- as feedstock (source?). Forest
residues are generated as by-products of conventional forest harvesting
operations, and include the tree tops, branches and non-merchantable
logs. These residues are typically left behind in slash piles, which are
burned during winter to mitigate against the risk of wildfires. Canada
has a developed forestry sector that produces approximately 20 million
t\[oven dried\]yr^-1^ of forest residues (Dymond et al. 2010).

Despite the potential to produce biofuels from forest residues, such use
is limited to a few highly subsidized pilot projects. Indeed, several
operational and economic challenges hinder the utilization of forest
residues. These challenges are related to: (1) capital investments in
biorefineries, (2) geographical distance between feedstock and markets,
and (3) concerns surrounding feedstock availability (Shabani et al.
2013). The purpose of our study is to explore concerns surrounding this
third challenge -- feedstock availability -- by evaluating annual
variability in commercial stemwood harvest in the province of Alberta,
Canada.

A number of studies have shown that collection and transportation costs
are arguably the biggest constraint to widespread use of forest residues
in bioenergy systems (Ralevic et al 2003, Wood and Layzell 2003, Kumar
et al. 2003, Rummer 2007, Aulakh 2008). Availability of forest residues
for a biorefinery largely depend on commercial harvesting operations and
the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products could have a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in forest
harvesting activities (De Avillez 2014, Spelter 2009). This relationship
between volatile sawlog harvests and subsequent residue variability is
explored in detail by Niquidet and Friesen (2014), who model forest
residue supply in Alberta as a direct function of lumber prices. Given
that biomass feedstock accounts for 40 to 60% of a biorefinery's total
costs (Stephen et al. 2012), reliable feedstock supplies and costs over
the life of the biorefinery are important factors to consider when
deciding where to locate the biorefinery.

@@

Although not related to a biorefinery siting decision, some studies have
considered variable feedstock supplies when investigating whether and
when to expand the production of bioenergy from burning residues
(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
Moiseyev et al. 2011). Other studies have explored the context of
variable supplies when making decisions about capital investments in
forest residue-based biorefineries (Cambero et al 2015, Chen and Fay
2011, Papapostolou and Kondili 2011). Variable feedstock supplies are
also shown to affect capital investment decisions related to: investing
in a residue preprocessing facility (Chen and Fay 2011), types of
storage infrastructure for biorefinery feedstocks (Papapostolou and
Kondili 2011), as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities (Cambero et al.
2015).

There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnston et al (2012) for
a review of these studies -- however, we are aware of no study that
considered stemwood harvest variability and its subsequent effect on
forest residue variability. Therefore, our study contributes to the
literature by considering variability in forest residue feedstock
supplies. More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a biorefinery.

We consider co-locating the biorefinery adjacent to either a sawmill or
a pulpmill. Co-locating with existing facilities and making use of
established infrastructure and transportation systems can reduce many of
the costs associated with establishing a new biorefinery (Browne et al.,
2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
2007). The decision to co-locate a biorefinery near a sawmill or
pulpmill should be informed by how stable the respective residue
supplies are. The facility with a more stable stemwood harvest would
generate a more stable supply of forest residues that could be collected
from the nearby harvested areas.

@@

**OTHER POTENTIALLY USEFUL TIDBITS...**

The need to go outside its normal delivery zone in bad years is similar
to a scenario in the agriculture residue feedstock supply literature,
called the the "derisked" scenario, whereby a buffer area outside the
normal supply zone is contracted to provide residues during years when
residue supply is low (Golecha and Gan 2016).

@@

Modern biorefineries can expect to breakeven financially with delivered
biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
(Steeper 2021) . Studies have shown that forest policy reforms could
reduce delivered forest residue costs by incenting plantations of
fast-growing tree species on lands close to pulpmill locations
(Shooshtarian et al. 2021, Anderson et al. 2012). Other important
factors will be advances in biorefinery technology and increases in
biofuel prices. In Canada, governments impact biofuel prices by setting
renewable fuel mandates and paying subsidies for biofuel production
(Campbell et al., 2016). The continuation of these mandates and
subsidies will likely be an important factor for the emergence of a
second-generation ("advanced") biofuel sector. Also, the Government of
Canada\'s Clean Fuel Standard, which is proposed to come into force in
2022 (Government of Canada, 2020), is expected to improve the economics
of advanced biofuel production.
`

@@

The NFIS also provides a web-based individual tree biomass calculator
@NFIS2021.  It is used to calculate stem wood, stem bark, branch, and
foliage biomass for a tree based on province, ecozone, species,
diameter at breast height (DBH) and height.  It is possible to upload
a file representing many trees with different combinations of these
input parameters and download a file containing the calculated results
for all the trees.

@@

This equates to `r sprintf("%5.2f",odt_trailer)` Mg
of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
according to Miles and Smith @Miles2009a.

>>>>

---
title: Spatial and temporal variability in economic availability of residual biomass
  from timber harvest in west-central Alberta, Canada (1990--2015)
author:
- Glen W. Armstrong:
    institute: rr
    correspondence: 'yes'
    email: gwa@ualberta.ca
- Jay A. Anderson:
    email: jaya@ualberta.ca
    institute: rees
- Alex W. McPhee:
    email: awmcphee@ualberta.ca
    institute: rees
institute:
- rr: Department of Renewable Resources, University of Alberta, Edmonton AB  T6G 2H1,
    Canada
- rees: Department of Resource Economics and Environmental Sociology, University of
    Alberta, Edmonton AB  T6G 2H1, Canada
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    includes:
      in_header: preamble.tex
    number_sections: yes
    latex_engine: xelatex
    keep_tex: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  bookdown::pdf_document2: default
  html_document:
    number_sections: yes
    df_print: paged
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  word_document:
    number_sections: yes
    pandoc_args:
    - --lua-filter=scholarly-metadata.lua
    - --lua-filter=author-info-blocks.lua
  bookdown::html_document2: default
header-includes:
- \usepackage[left]{lineno}
- \usepackage{setspace}
- \linenumbers
geometry: left=1.5in,right=0.5in,top=1in,bottom=1in
csl: biomass-and-bioenergy
bibliography: historicalResidueSupply.bib
---

@@

```{r libraries, echo=FALSE, message=FALSE}
#setwd("G:/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/g/Shared drives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
#setwd("/mnt/chromeos/GoogleDrive/SharedDrives/FES_Data_Sharing_Forestry/Manuscripts/histressupp/analysis")
library(tidyverse)
library(ggthemes)
library(gridExtra) 
library(viridis)
library(skimr)
library(scales)
library(DescTools)
library(zoo)
library(ezknitr)
library(magick)
library(kableExtra)
library(gt)
library(glue)
library(quantmod)
library(bookdown)
```

@@

Biofuels are becoming an increasingly important source of energy because
of their potential to reduce dependency on fossil fuels, thereby
reducing greenhouse gas emissions (IEA Bioenergy 2013). Moreover, second
generation biofuel provides the benefit of avoiding competition with
food production by using lignocellulosic biomass -- often in the form of
agriculture and forest residues -- as feedstock (source?). Forest
residues are generated as by-products of conventional forest harvesting
operations, and include the tree tops, branches and non-merchantable
logs. These residues are typically left behind in slash piles, which are
burned during winter to mitigate against the risk of wildfires. Canada
has a developed forestry sector that produces approximately 20 million
oven-dry metric tons  of forest residues (Dymond et al. 2010) @Dymond2010.

Despite the potential to produce biofuels from forest residues, such use
is limited to a few highly subsidized pilot projects. Indeed, several
operational and economic challenges hinder the utilization of forest
residues. These challenges are related to: (1) capital investments in
biorefineries, (2) geographical distance between feedstock and markets,
and (3) concerns surrounding feedstock availability (Shabani et al.
2013) @Shabani2013. The purpose of our study is to explore concerns surrounding this
third challenge -- feedstock availability -- by evaluating annual
variability in commercial stemwood harvest in the province of Alberta,
Canada.

A number of studies have shown that collection and transportation
costs are arguably the biggest constraint to widespread use of forest
residues in bioenergy systems (Ralevic et al 2003, Wood and Layzell
2003, Kumar et al. 2003, Rummer 2007, Aulakh 2008)
[@Kumar2003,@Rummer2007,@Aulakh2008]. Availability of forest residues
for a biorefinery largely depend on commercial harvesting operations
and the production of traditional forest products. Fluctuations in the
volume of stemwood harvested for forest products could have a direct
impact on residues available for biofuels. Major factors, such as the
2008 housing crisis and the ongoing softwood lumber dispute with the
United States have resulted in sawmill closures and reductions in
forest harvesting activities (de Avillez 2014, Spelter 2009)
@deAvillez2014 @Spelter2009. This relationship between volatile sawlog
harvests and subsequent residue variability is explored in detail by
Niquidet and Friesen (2014) @Niquidet2014, who model forest residue
supply in Alberta as a direct function of lumber prices. Given that
biomass feedstock accounts for 40 to 60% of a biorefinery's total
costs (Stephen et al. 2010) @Stephen2010, reliable feedstock supplies
and costs over the life of the biorefinery are important factors to
consider when deciding where to locate the biorefinery.

@@

Although not related to a biorefinery siting decision, some studies
have considered variable feedstock supplies when investigating whether
and when to expand the production of bioenergy from burning residues
(Bolkesjo et al. 2006; Buongiorno et al. 2011; Galik et al. 2009;
Moiseyev et al. 2011) @Bolkesjo2006 @Buongiorno2011 @Galik2009. Other
studies have explored the context of variable supplies when making
decisions about capital investments in forest residue-based
biorefineries (Cambero et al 2015, Chen and Fay 2011, Papapostolou and
Kondili 2011) @Cambero2015, @Chen2012 @Papapostolou2011. Variable
feedstock supplies are also shown to affect capital investment
decisions related to: investing in a residue preprocessing facility
(Chen and Fay 2011) @Chen2012, types of storage infrastructure for
biorefinery feedstocks (Papapostolou and Kondili 2011)
@Papapostolou2011, as well as whether to invest in biofuels versus
bioenergy, and when to invest in processing facilities (Cambero et al.
2015) @Cambero2015.

There has been a number of studies conducted regarding siting decisions
for forest residue-based biorefineries -- see Johnson et al (2012) @Johnson2012  for
a review of these studies -- however, we are aware of no study that
considered stemwood harvest variability and its subsequent effect on
forest residue variability. Therefore, our study contributes to the
literature by considering variability in forest residue feedstock
supplies. More specifically, we analyze how this variability affects the
cost of delivering residue, and then use these cost estimates to inform
the decision on where to locate a biorefinery.

We consider co-locating the biorefinery adjacent to either a sawmill or
a pulpmill. Co-locating with existing facilities and making use of
established infrastructure and transportation systems can reduce many of
the costs associated with establishing a new biorefinery (Browne et al.,
2012; Saddler et al., 2012; Benjamin et al., 2009; Towers, et al.,
2007). @Browne2012 @Saddler2012 @Benjamin2009 @Towers2007 The decision to co-locate a biorefinery near a sawmill or
pulpmill should be informed by how stable the respective residue
supplies are. The facility with a more stable stemwood harvest would
generate a more stable supply of forest residues that could be collected
from the nearby harvested areas.

@@

**OTHER POTENTIALLY USEFUL TIDBITS..**

The need to go outside its normal delivery zone in bad years is similar
to a scenario in the agriculture residue feedstock supply literature,
called the the "derisked" scenario, whereby a buffer area outside the
normal supply zone is contracted to provide residues during years when
residue supply is low (Golecha and Gan 2016) @Golecha2016.

@@

Modern biorefineries can expect to breakeven financially with delivered
biomass costs of approximately \$80 (US\$60) t\[oven dried\]^-1^
(Steeper 2021) @Steeper2021 . Studies have shown that forest policy reforms could
reduce delivered forest residue costs by incenting plantations of
fast-growing tree species on lands close to pulpmill locations
(Shooshtarian et al. 2021 @Shooshtarian2021, Anderson et al. 2012 @Anderson2012). Other important
factors will be advances in biorefinery technology and increases in
biofuel prices. In Canada, governments impact biofuel prices by setting
renewable fuel mandates and paying subsidies for biofuel production
(Campbell et al., 2016). The continuation of these mandates and
subsidies will likely be an important factor for the emergence of a
second-generation ("advanced") biofuel sector. Also, the Government of
Canada\'s Clean Fuel Standard, which is proposed to come into force in
2022 (Government of Canada, 2020), is expected to improve the economics
of advanced biofuel production.
`

@@

The NFIS also provides a web-based individual tree biomass calculator
@NFIS2021a.  It is used to calculate stem wood, stem bark, branch, and
foliage biomass for a tree based on province, ecozone, species,
diameter at breast height (DBH) and height.  It is possible to upload
a file representing many trees with different combinations of these
input parameters and download a file containing the calculated results
for all the trees.

@@

This equates to `r sprintf("%5.2f",odt_trailer)` Mg
of biomass assuming a specific gravity of `r sprintf("%5.2f",specific_gravity)` 
which is the specific gravity of lodgepole pine wood and bark on a green volume basis 
according to Miles and Smith @Miles2009.
